Factors affecting reliablity of state-space age-structured assessment models

### 2025-04-08 Project_0/manuscript/manuscript.Rmd
internal review for project_0 paper

$^1$corresponding author: timothy.j.miller@noaa.gov\
$^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
$^3$Woods Hole Oceanographic Institution\
$^4$SMAST\
$^5$Institute of Marine Research\
$^6$Mid-Atlantic Fisheries Management Council\
$^7$Rutgers University\

@@

@@

<!-- note CJFAS abstract word limit 175, so I have drafted a shorter summary here -->

State-space assessment models, which are increasingly being used for management of commercially important fish stocks, can include process errors in many ways, and an evaluation of the statistical reliability of the estimates of management output and best practices of selection among alternative process error configurations are needed. We simulated 72 operating models with varying fishing pressure history, observation error error, and process error magnitude, correlation, and source (recruitment, survival, fishery selectivity, catchability, and natural mortality). We fit estimating models with different assumptions on the process error source, whether median natural mortality was estimated, and whether a stock-recruit relationship was estimated. Estimating models that assumed the correct process error source and median natural mortality rate was known, and did not estimate a stock recruit relationship generally had high probability of convergence and low bias in terminal year SSB estimation. Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error. Stock-recruitment parameters were not reliably estimated in any scenarios, but the least accurate estimates occured under ideal situations of low observation error, low recruitment variability, and temporal contrast in fishing pressure. However, estimation of median natural mortality rate was reliable for many OMs (other than those with survival process errors) with temporal contrast in fishing pressure regardless of the estimated process error type. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types. Retrospective patterns for spawning stock biomass and fishing mortality were generally weak, but those for recruitment were sizable when observation error was high, even with the correct process error assumptions.

@@

Application of state-space models in fisheries stock assessment and management has expanded dramatically within International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21]. State-space models latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error. Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effects parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15]. Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches. These latter approaches are considered best practice  and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].

@@

Despite mounting efforts, several limitations remain. First, confounding processes that can be treated as random effects in the model were not thoroughly examined or tested within a simulation-estimation framework. Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24]. In particular, the effects of observation error and underlying fishing history were not fully isolated from their study design, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23. Third, explicitly modeling stock-recruitment relationships as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22] but can be challenging in state-space models. Evidence suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09]. In particular, @knape08 demonstrated that stronger density dependence becomes increasingly difficult to estimate in the presence of observation error. Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision before drawing ecological conclusions [@augeretal16].

In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated. Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21]. We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a stock-recruit relationship. We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available as a github repository. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). This package has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.

We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q). For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.

In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time. We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated. For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs. Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM. Details of each of the operating and EMs are described below.

We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for operating and EMs [@lietalinreview]. Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.

@@

The population consists of 10 age classes: ages 1 to 10+ and we assume spawning occurs each year 1/4 of the way through the year. The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).

@@

We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all OMs. All post-recruit productivity components are constant in the NAA and survey catchability process error OMs. Therefore steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21]. We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the
\[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\]
Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom right). For OMs without process errors on natural mortality we assumed the rate was assumed 0.2. For OMs with process errors on natural mortality the median natural mortality rate was 0.2.

We used two fishing scenarios for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years and fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$). In the second scenario, the stock is fished at \Fmsy for the entire time period. The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from [@wiedenmannetal19]. @legaultetal23 also used similar approaches to defining fishing mortality histories for OMs.

We specified initial population abundance at age at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories. This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.

For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the median of the time series process for M. For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other OMs which corresponds to Fmsy at the mean selectivity parameters.

@@

We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left). This selectivity is was used to define \Fmsy for the Beverton-Holt stock recruitment parameters above. We assumed a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations for the fleet.

@@

Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).  Catchability of both surveys are assumed to be 0.1.  Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations.

@@

Standard deviation for log-aggregate catch was 0.1. There were two levels of observation error variance for indices and age composition for both indices and fleet catch. A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5. For all EMs, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.

@@

For operating models with random effects on recruitment and(or) survival (R, R+S) we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$. The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).

### Operating models with random effects on natural mortality {-}

All R+M OMs treat natural mortality constant across age, but with annually varying random effects. WHAM treats natural mortality as a log-transformed parameter 
$$
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
$$
that is a linear combination of a mean log-natural mortality parameter that is constant across ages $\mu_{M} = \log(0.2)$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$. Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, R+M). The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_M \in \{0,0.9\}$. The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

MORE SPECIFICS about correlation  of random effects? Both selectivity pars? just correlated by year?
WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
$$
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
$$
that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$ where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user. All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and EMs. The marginal standard deviations we assumed for logit scale random effects were $\sigma_s \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_s \in \{0,0.9\}$.  Like R+M OMs,  the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+Sel OMs (Table \ref{sel_om_table}).

@@

Like selectivity parameters, WHAM treats catchability for an index $i$ as a logit-transformed parameter 
$$
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
$$
that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user. Here we assume bounds of 0 and 1000 for all operating and EMs. For operating and EMs with process errors on catchability, the temporal variation is only assumed for the first index. The marginal standard deviations we assumed for logit scale random effects were $\sigma_i \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_i \in \{0,0.9\}$.  Like R+M and R+Sel OMs, the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+q OMs (Table \ref{q_om_table}).

@@

For each data set simulated from an OM 20 EMs were fit. A total of 32 different EMs were fit across OMs where the subset of 20 depended on the source of process error in the OM (Table \ref{em_table}). The EMs make different assumptions about the source of process error (R+S, R+M, R+Sel, R+q) and whether there is temporal autocorrelation or not, and whether a Beverton-Holt stock recruit function is estimated and the natural mortality rate ($\mu_M$, the constant or mean on log scale for R+M EMs) is estimated. For simplicity we refer to the derived estimate $e^{\mu_M}$ as the median natural mortality rate regardless of whether natural mortality random effects are estimated in the EM. The first 20 EMs in Table \ref{em_table} were fit to simulate data sets from R and R+S OMs. EMs 5 to 24 in Table \ref{em_table} were fit to simulate data sets from R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were fit to simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32 in Table \ref{em_table} were fit to simulate data sets from R+q OMs. The maturity at age, weight at age for catch and SSB, and observation error variance of aggregate catch and indices were all assumed known at the true values.

@@

The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets.
There are various ways to assess convergence of the fit [e.g., @carvalhoetal21], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted. However, we also provided results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects. This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences.

<!--
probability of convergence as measured by all gradient values at the optimized negative log-likelihood that were less than $1\times10^{-6}$ in absolute value. We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].
-->

@@

We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given operating model, we compared AIC for EMs that all made the same assumptions about median natural mortality (known or estimated) and stock recruitment model (Beverton-Holt or none).

We also estimated the probability of correctly selecting between models with Beverton-Holt stock recruit function assumed and models without the stock-recruit function (null model). We made these comparisons between models that otherwise assume the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known. Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimability of stock-recruitment parameters [@magnussonhilborn07;@connetal10]. Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors (including contrast in fishing pressure) indicated generally poor performance of AIC. Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) since simulations with realized SSB producing low and high recruitments would have larger variation in realized SSB.

All results only condition on whether all of the compared estimating models completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization can correctly determine the the correct likelihood that would indicate poor convergence because variance parameters may be at the lower bound of zero correctly for models that assume the incorrect process error structure.

@@

For a given model attribute we calculated the relative error 
$$
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$. We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination. We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21. We present results for terminal years SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate. Results for terminal year fishing mortality are strongly negatively correlated with those for SSB and provided in the supplementary materials.

Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects. In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.

@@

When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means its the correct process error .

>>>>

$^1$corresponding author: timothy.j.miller@noaa.gov\
$^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
$^3$Woods Hole Oceanographic Institution\
$^4$SMAST\
$^5$Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\
$^6$Institute of Marine Research, Nye Flødevigveien 20, 4817 His, Norway\
$^7$Department of Ecology, Evolution, and Natural Resources. Rutgers University\

@@

**keywords:** stock assessment, state-space, model selection, bias, convergence, retrospective patterns

@@

State-space models are increasingly used for stock assessment, and evaluations of their statistical reliability and best practices for selecting among process error configurations are needed. We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, survival, selectivity, catchability, and/or natural mortality. We fit estimating models with different assumptions on the process error source and whether median natural mortality or a stock-recruit relationship were estimated. Estimating models without a stock-recruit relationship that assumed the correct process error source and median natural mortality had high convergence rates and low bias. Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, as well as larger magnitude process errors of other types. Retrospective patterns were generally small but were sizable for recruitment when observation error was high. These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment

@@

Application of state-space models in fisheries stock assessment and management has expanded dramatically within International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21]. State-space models latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error. Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effects parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15]. Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches. These latter approaches are considered best practice and a recommended for the next generation of stock assessment models [@hoyleetal22; @punt23].

@@

Despite mounting efforts, several limitations remain. First, confounding processes that can be treated as random effects in the model were not thoroughly examined or tested within a simulation-estimation framework. Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24]. In particular, the effects of observation error and underlying fishing history have not been fully isolated in simulation study designs, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23. Third, explicitly modeling stock-recruit relationships (SRRs) as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22], but can reliability of inferences within integrated state-space age-structured models has not been evaluated. Evidence from other studies suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09]. In particular, @knape08 demonstrated that stronger density dependence becomes increasingly difficult to estimate in the presence of observation error. Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision for use in fisheries management [@augeretal16].

In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a SRR was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated. Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21]. We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a SRR. We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available via a github repository and is built on the Template Model Builder package [@kristensenetal16]. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous fish stocks in the Northeast United States (NEUS) [e.g., @nefsc22;@nefsc22a;@nefsc24].

We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q). We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects ($\epsilon_{a,y}$) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term. However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.  For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from NEUS assessments using WHAM in stocks NEUS.

In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time. We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated. Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM. Details of each of the OMs and EMs are described below.

We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview]. Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.

@@

The population consists of 10 age classes, ages 1 to 10+, with the last being a plus group that accumulates ages 10 and older. We assume spawning occurs annually 1/4 of the way through the year. The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left).

@@

We assumed a Beverton-Holt SRR with constant pre-recruit mortality parameters for all OMs. All post-recruit productivity components are constant in the apparent survival (R+S) and survey catchability (R+Sel) process error OMs. Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21]. We specified unfished recruitment equal to $e^{10}$ and $\Fmsy = F_{40\%} = 0.348$, which equates to a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ for the Beverton-Holt parameterization
\[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\]
(Figure \ref{om_inputs_fig}, bottom right). For OMs without process errors on natural mortality we fixed the rate at 0.2. For OMs with process errors on natural mortality, the median natural mortality rate was specified to be 0.2.

We used two fishing scenarios for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$). In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years). The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.

We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$. This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.

For OMs with time-varying random effects for M , steepness is not constant. However, we used the same $a$  and $b$ parameters as other OMs, which equates to a steepness and R0 at the median of the time series process for M. For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant, but since we use the same F history as other OMs, this corresponds to \Fmsy at the mean selectivity parameters.

@@

We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet ($a_{50} = 5$ and slope = 1; Figure \ref{om_inputs_fig}, bottom left). This selectivity was used to define \Fmsy for the Beverton-Holt SRR parameters above. We assumed a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations for the fleet.

@@

Two time series of fishery-independent surveys in numbers are generated for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).  Catchability of both surveys are assumed to be 0.1.  Like the fishing fleet, we assumed logistic selectivity for both indices ($a_{50} = 5$ and slope = 1) and a logistic-normal distribution with no correlation on the multivariate normal scale for the age-composition observations.

@@

The standard deviation for log-aggregate catch was 0.1. Two levels of observation error variance (high and low) were specified for indices and all age composition observations (both indices and catch). The low uncertainty specification assumed a standard deviation of 0.1 for both series of log-aggregate index observations, and the standard deviation of the logistic-normal for age composition observations was 0.3. In the high uncertainty specification, the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5. For all EMs, the standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.

@@

For operating models with random effects on recruitment and(or) apparent survival (R, R+S), we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$. The full factorial combination of these process error assumptions (2x3 levels) and scenarios for fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).

### Operating models with random effects on natural mortality {-}
All R+M OMs treat natural mortality as constant across age, but with annually varying random effects. WHAM treats natural mortality as a log-transformed parameter 
$$
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
$$
that is a linear combination of a mean log-natural mortality parameter that is constant across ages ($\mu_{M} = \log(0.2)$) and any annual random effects are marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$. Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, we denote these OMs as  R+M). The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_M \in \{0,0.9\}$. The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
$$
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
$$
that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$, where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user. All selectivity parameters ($a_{50}$ and slope parameters) were bounded by 0 and 10 for all OMs and EMs. The marginal standard deviations we assumed for logit scale random effects were $\sigma_s \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_s \in \{0,0.9\}$.  Like R+M OMs, the full factorial combination of these process error assumptions (2x2 levels) and scenarios described above for fishing history (2 levels) and observation error (2 levels) results in 16 different R+Sel OMs (Table \ref{sel_om_table}).

@@

Like selectivity parameters, WHAM treats catchability for an index $i$ as a logit-transformed parameter 
$$
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
$$
that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user. We assumed bounds of 0 and 1000 for all OMs and EMs. For all OMs and EMs with process errors on catchability, the temporal variation only applies to the first index, which could be interpreted as capturing some unmeasured seasonal process that affects availability to the survey. The marginal standard deviations we assumed for logit scale random effects were $\sigma_i \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_i \in \{0,0.9\}$.  Like R+M and R+Sel OMs, the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+q OMs (Table \ref{q_om_table}).

@@

For each of the data sets simulated from an OM, 20 EMs were fit. A total of 32 different EMs were fit across OMs where the subset of 20 depended on the source of process error in the OM (Table \ref{em_table}). The EMs have different assumptions about the source of process error (R+S, R+M, R+Sel, R+q) and whether or not 1) there is temporal autocorrelation, 2) a Beverton-Holt SRR is estimated, and 3) the natural mortality rate ($\mu_M$, the constant or mean on log scale for R+M EMs) is estimated. For simplicity we refer to the derived estimate $e^{\mu_M}$ as the median natural mortality rate regardless of whether natural mortality random effects are estimated in the EM.

Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets from each of the OM process error categories. For R and R+S OMs, fitted EMs had matching process error assumptions as well as R+Sel, R+M, and R+q assumptions without autocorrelation. Similarly, For other OM process error categories, we fit EMs with matching process error assumptions as well as other process error types without autocorrelation. The maturity at age, weight at age for catch and SSB, and observation error variance of aggregate catch and indices were all assumed known at the true values. However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs.

@@

### Convergence

The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets. There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted. We also provide results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects. This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences. We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].

@@

We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given operating model, we compared AIC for EMs that all made the same assumptions about median natural mortality (known or estimated) and SRR (Beverton-Holt or none).

We also estimated the probability of correctly selecting between EMs with Beverton-Holt SRR assumed and models without the SRR (null model). We made these comparisons between models that otherwise assumed the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known. Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10]. Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors (including contrast in fishing pressure) indicated generally poor performance of AIC. Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) since simulations with realized SSB producing low and high recruitments would have larger variation in realized SSB.

All model selection results condition on whether all of the compared estimating models completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero. Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated. All other measures of reliability described below (bias and Mohn's $\rho$) use these same criteria for inclusion of EM fits in the summarized results.

@@

For a given model attribute we calculated the relative error 
$$
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$. We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination. We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21. We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate. Results for terminal year fishing mortality are were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.

<!--
Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of assessment model output whether or not the model was able to estimate non-zero random effects. In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.
-->

@@

When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means its the correct process error .
-->
